---
title: "lme4 tutorial"
output: html_notebook
---


```{r libraries, include=FALSE}
# lme4 tutorial
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom)
library(ggfortify)
library(stargazer)
```

```{r}
# sleepstudy dataset contains reaction times after sleep deprivation
?sleepstudy

# First, check the data as the data was independent
ggplot(data = sleepstudy) +
    aes(x = Days, y = Reaction) +
    geom_point() +
    geom_smooth(method = "lm")

sleep.lm <- lm(Reaction ~ Days, data = sleepstudy)
# Check predictor statistics
summary(sleep.lm)
tidy(sleep.lm)
# Check model fit statistics
glance(sleep.lm)
```

```{r}
# Residual diagnostic plots
autoplot(sleep.lm, which = 1:6)
```

# Check if the mixed-model does more sense
First try a random intercept and slope model, with all fixed effects included. This backward elimination protocol was suggested by Zuur(2009)

The lme4 syntax is very similar to the lm syntax. You can add random terms with + (<slope>|<intercept>). If you set the slope to 1, it means it remains fixed

For a repeated measures design, the random intercept should be the subject (id)
Let's check if the random intercept (but not slope model) is a better fit than the fixed

```{r}
# I want to show the fixed slope, random intercept model
# To do this, I fit an lm to each subject individually. To get the fixed intercept, I take the average intercept in all models.
subject_is <- 
    sleepstudy %>% 
    group_by(Subject) %>% 
    do(lm = lm(Reaction ~ Days, data = .) %>% tidy()) %>% 
    unnest() %>% 
    select(Subject, term, estimate) %>% 
    mutate(term = case_when(term == "(Intercept)" ~ "intercept",
                            term == "Days" ~ "slope",
                            TRUE ~ NA_character_)) %>% 
    spread(term, estimate) %>% 
    mutate(fixed_slope = mean(slope))

# Plot the random intercept model
sleepstudy %>% 
    left_join(subject_is) %>% 
    ggplot() +
        aes(x = Days, y = Reaction) +
        geom_abline(aes(intercept = intercept, slope = fixed_slope), color = "blue", size = 1.2) +
        geom_point() +
        facet_wrap(~Subject)

# Fit the random intercept model
sleep.i <- lmer(Reaction ~ Days + (1|Subject), data = sleepstudy)

summary(sleep.i)
glance(sleep.i)

```
We can see that the model looks better, but the predicted values are sometimes quite off, e.g. with particpant 308 and 335.

# Compare the two models with and without random intercept
The one with the random intercept has a much better fit, although it is more compex.
lme4 by default uses REML for fitting the model, but when we are comparing models, it automatically refits the models in ML, so they are comparable.
```{r}
anova(sleep.i, sleep.lm)
```

# Let's check if we need random slope too, or random intercept is enough
It turns out that the random intercept and slope model is the better fit, so it should be kept as final
```{r}
# First, visualize how the data would look like with random intercept and slope. With ggplot, it is easy, beacause the geom_smooth(method = "lm") function does this automatically, we don't have to use our data from the earlier chunk.
ggplot(sleepstudy) +
    aes(x = Days, y = Reaction) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(~Subject)

sleep.is <- lmer(Reaction ~ Days + (Days|Subject), data = sleepstudy)

summary(sleep.is)
glance(sleep.is)
anova(sleep.i, sleep.is)
```

# Residual diadnostics of the linear mixed-model
Sadly, there autoplot is not capable to make the residual diagnostic plots, but we can do it easily using the results of the augment() function.
Despite the three outlier data points, the residuals look rather normal. 
We should accept this model as the final one.

```{r}
diag_data <- augment(sleep.is, sleepstudy) %>% 
    mutate(.stdresid = scale(.resid) %>% as.numeric())

ggplot(diag_data) +
    aes(x = .fitted, y = .resid) +
    geom_point() +
    geom_smooth()

ggplot(diag_data) +
    aes(x = .fitted, y = sqrt(abs(.stdresid))) +
    geom_point() +
    geom_smooth()

ggplot(diag_data) +
    aes(sample = .stdresid) +
    geom_qq_line(color = "red") +
    geom_qq()

ggplot(diag_data) +
    aes(x = .stdresid) +
    geom_histogram(bins = 30)


```

# Print model comparison using stargazer
We need to refit the mixed-models using the lme4 version of lmer, as stargazer cannot read the lmerTest::lmer() object.

```{r, results = "asis"}
sleep.i.final <- lme4::lmer(Reaction ~ Days + (1|Subject), data = sleepstudy)
sleep.is.final <- lme4::lmer(Reaction ~ Days + (Days|Subject), data = sleepstudy)

stargazer(sleep.lm, sleep.i.final, sleep.is.final, type = "html")
```







